---
title: "Classifying outage events by length"
output: html_notebook
---

This notebook reads `ca_pct_household_oop_new_def.csv` and separate outage events at the county level of 2019 in California into the following categories:

- Major: < 1 day
- Severe: [1,3) days
- Extreme: 3+ days

## Read libraries
```{r}
library(tidyverse) # ggplot2, dplyr, tidyr, readr, purrr, tibble
library(magrittr) # pipes
library(sf) # spatial data handling
library(rmarkdown)
library(here)
library(ggExtra)
library(rgeoda)
```

## Clean work environment
```{r}
rm(list = ls())
```

## Read file with power outage events

```{r}
df_outage <- read_csv(here("result","power-outage-events",
                           "ca_pct_household_oop_new_def.csv"))
head(df_outage)
```

## Classify events by `Duration`

- Remove very short outages (< 60 min)

```{r}
df_outage <- df_outage %>%
    filter(Duration >= 60)
```

- Get major events (< 24 hours)

```{r}
major_outage <- df_outage %>%
    filter(Duration < 60*24) %>%
    group_by(County) %>%
    summarise(Count = n(),
              SumMaxCust = max(SegCustomerOut))
```

- Get severe events (1-3 days)

```{r}
severe_outage <- df_outage %>%
    filter(Duration >= 60*24 & Duration <= 60*24*3) %>%
    group_by(County) %>%
    summarise(Count = n(),
              SumMaxCust = max(SegCustomerOut))
```

- Get extreme events (1-3 days)

```{r}
extreme_outage <- df_outage %>%
    filter(Duration > 60*24*3) %>%
    group_by(County) %>%
    summarise(Count = n(),
              SumMaxCust = max(SegCustomerOut))
```

## Create a Shapefile for each outage type

- Read a shell Shapefile for county boundaries

```{r}
ca_cnty <- read_sf(here("data_private","ca_county_basic_geography.shp"))
head(ca_cnty)
```

- Add inner join between the shell Shapefile and df for outages

```{r}
major_outage_shp <- ca_cnty %>%
    left_join(major_outage, by = c("NAME" = "County")) %>%
    replace_na(list(Count = 0, SumMaxCust = 0))
```

```{r}
severe_outage_shp <- ca_cnty %>%
    left_join(severe_outage, by = c("NAME" = "County")) %>%
    replace_na(list(Count = 0, SumMaxCust = 0))
```

```{r}
extreme_outage_shp <- ca_cnty %>%
    left_join(extreme_outage, by = c("NAME" = "County")) %>%
    replace_na(list(Count = 0, SumMaxCust = 0))
```

## Write results as Shapefiles

```{r}
st_write(major_outage_shp, here("result","ca_county_power_disruption_major.shp"), 
         delete_layer = TRUE)

st_write(severe_outage_shp, here("result","ca_county_power_disruption_severe.shp"), 
         delete_layer = TRUE)

st_write(extreme_outage_shp, here("result","ca_county_power_disruption_extreme.shp"), 
         delete_layer = TRUE)
```