---
title: "Create a bar plot showing frequencies of dservice disruptions in California (2019)"
author: Shenyue Jia
output: html_notebook
---

This notebook plots a bar plot for power outage events at different lengths for Exhibit 2 in the manuscript.

```{r setup, warning=FALSE}
# loading libraries
library(tidyverse)
library(zoo)
library(foreign)
library(here)
library(purrr)
library(lubridate)
library(ggthemes)
library(viridis)
library(classInt)
library(hrbrthemes)
library(ggridges)
```

```{r}
rm(list = ls())
```

# Read California state-level aggregated data showing power outages

- Get county-level power outage counts

```{r}
county_outage_df <- read_csv(here("result",
                                  "ca_pct_household_oop_new_def.csv"))

```

- Add a Month column
```{r message=FALSE, warning=FALSE}

county_outage_df <- county_outage_df %>%
    mutate(Month = factor(month(Start,label = TRUE)))

```

# Count power outage events at different lengths in CA

- Create a data frame that contains a type field for disruptions based on duration
    - 1-8 hrs
    - 9-16 hrs
    - 17-24 hrs
    - 24+ hrs

```{r message=FALSE, warning=FALSE}
df_plot <- bind_rows(
                    county_outage_df %>%
                      filter(Duration >=60 & Duration<=8*60) %>%
                      mutate(Type = "1-8 hrs"),
                    county_outage_df %>%
                      filter(Duration >8*60 & Duration<=16*60) %>%
                      mutate(Type = "9-16 hrs"),
                    county_outage_df %>%
                      filter(Duration >16*60 & Duration<=24*60) %>%
                      mutate(Type = "17-24 hrs"),
                    county_outage_df %>%
                      filter(Duration >24*60) %>%
                      mutate(Type = "24+ hrs"))
df_plot$Type <- as.factor(df_plot$Type)
df_plot$Month <- as.factor(df_plot$Month)
```

# Plot results as stacked bar plot

- Prepare a dataframe for bar plotting
```{r message=FALSE, warning=FALSE}

df_plot$Type <- factor(df_plot$Type,
                                 levels=c("1-8 hrs","9-16 hrs","17-24 hrs","24+ hrs"))

# extrafont::loadfonts()
df_plot_count <- df_plot %>%
  group_by(Month,Type) %>%
  summarise(Count = n()) %>%
  complete(Month,Type,fill=list(Count=0)) %>%
  mutate(label = ifelse(Count==0,NA,Count))

df_plot_count_first <- df_plot_count %>%
  group_by(Month, Type) %>%
  summarise(Month = first(Month),
            Type = first(Type),
            Count = first(Count),
            label = first(label))
```

- Generate bar plot

```{r message=FALSE, warning=FALSE}

df_plot_count_first %>%
  # ggplot(aes(x = Month,y=Count,fill=Type,group=Type)) +
  ggplot(aes(x = Month,y=Count,fill=Type)) +
  geom_col(position = position_stack(),width=0.5)+
  scale_fill_manual(values=c("#CABED0", "#BC7C8F", "#AE3A4E","#330000"))+
  scale_x_discrete(breaks = c("Jan","Feb","Mar","Apr","May","Jun","Jul",
                              "Aug","Sep","Oct","Nov","Dec"),
                   labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul",
                              "Aug","Sep","Oct","Nov","Dec 2019"))+
  scale_y_continuous(labels=function(x) format(x, big.mark = ".", scientific = FALSE))+
  guides(fill=guide_legend(title=element_blank()))+
  # facet_wrap(~Type,ncol =1)+
  # theme_ipsum() +
  theme_bw()+
    theme(
      legend.position="bottom",
      panel.spacing = unit(c(0,0,0,0), "mm"),
      plot.margin = unit(c(0,0,0,0), "mm"),
      axis.title.x = element_blank()
    )+
  # ggtitle("b) Major, severe, and extreme disruptions by starting months")+
  labs(y="Monthly Frequency of power disruptions") -> plot
plot
```

- Export the plot
```{r}
ggsave(
  here("plot","02_barplot_outage_count_notrans_stacked.pdf"),
  p3,
  # device = cairo_pdf,
  width = 6,
  height = 3,
  scale = 1
)
```