---
title: "Relationship between extreme power outage events and low-income populations"
author: Shenyue Jia
output: html_notebook
---

This notebook creates a scatterplot showing the relationship between the extreme power outage events in California, 2019 and low-income population (defined as 200% below federal poverty line). Extreme power outage events are defined as a power outage lasts for more than 24 hours. Low-income population is derived from [HRSA UDPMapper](https://udsmapper.org/).

```{r setup, include=FALSE}
library(rstudioapi)
library(tidyverse) # ggplot2, dplyr, tidyr, readr, purrr, tibble
library(magrittr) # pipes
library(lintr) # code linting
library(sf) # spatial data handling
library(viridis) # viridis color scale
library(cowplot) # stack ggplots
library(rmarkdown)
library(here)
library(hrbrthemes)
library(ggExtra)
```

```{r}
rm(list = ls())
```

# Data preparation

- Prepare California county-level extreme outages and cumulatively % of affected households

```{r}
ca_outage <- read_csv(here("result", "ca_pct_household_oop_new_def.csv"))
ca_outage_24plushrs <- ca_outage %>%
  filter(Duration>24*60) %>%
  group_by(County) %>%
  summarise(SumPctHU = sum(PctHU)) %>%
  complete(County,fill=list(SumPctHU=0))

```

- Prepare California county-level data from a pre-pared Shapefile
    - County-level metrics from census
    - County-level metrics from HHS emPOWER platform on DME users and Medicare beneficiaries
    - County-level metrics from HRSA UDSMapper on healthcare resources availability

```{r}

ca_geo <- read_sf(here("data_private",
                       "ca_county_census_empower_UDSMapper_combined.shp"))

```

# Clean variables and prepare normalized variables

- Convert selected variables from `ca_geo` to normalized versions

```{r}
ca_geo_sub <- ca_geo %>%
  mutate(pct_low_income = round((1-pene_low_i)*100,2)) %>%
  select(STATEFP,COUNTYFP,GEOID,NAME,x_map,y_map,
         SumMaxCust,pct_low_income,TOTHU_CY_1) %>%
  mutate(avg_outage_customer = SumMaxCust/TOTHU_CY_1)

```

# Visualization

- Prepare a function to create beautiful breaks

```{r}
  calculate_class <- function(var,nclass,roundto){
    class_brks <- classInt::classIntervals(var, n = nclass,intervalClosure = "left",
                                 style = "equal")
    class_brks <- class_brks$brks
    class <- plyr::round_any(class_brks,roundto)
    
    return(class)
  }
```

- Annotate counties with both metrics high

```{r}
ca_geo_sub$label <- NA

ca_geo_sub %<>%
  mutate(label = ifelse((avg_outage_customer>0.5 & 
           pct_low_income>25),NAME,NA))

```

- Add fill color for labels by region
```{r}
ca_region <- 
  read_csv(here("data_private","california_regions.csv")) %>%
    mutate(NAME = substr(COUNTY, 1, nchar(COUNTY)-nchar(" County")))
```

- Setup colors for regions
```{r}
reg_hex_code <- c("#FCB0AF","#F8F892","#C2E5A4","#66B460","#EB984E","#AED6F1")
reg_hex_code_2 <- c("#C2E5A4","#66B460","#FCB0AF","#F8F892","#EB984E","#AED6F1")

```

Join the region color to ca_geo_sub
```{r}
ca_geo_sub <- left_join(ca_geo_sub, ca_region,by="NAME") %>%
  mutate(bold_label = ifelse(label %in% c("Plumas","Shasta","Butte","Tehama","Sierra"),
                       label,NA),
    label = ifelse(label %in% c("Plumas","Shasta","Butte","Tehama","Sierra"),
                       NA,label))
```

- Create a scatterplot with selected counties labeled
```{r}
scatterplot <- 
            ggplot(data = ca_geo_sub,
                aes(
                    x = avg_outage_customer,
                    y = pct_low_income
                )
            ) + 
              theme_ipsum()+
              theme(panel.border = element_rect(linetype = "solid", fill = NA),
                    axis.text.x = element_text(size=10),
                    axis.text.y = element_text(size=10),
                    axis.title.x = element_text(size=12),
                    axis.title.y = element_text(size=12))+
              geom_point(alpha = .6, aes(size=TOTHU_CY_1)) +
             ggrepel::geom_label_repel(
                aes(label = label,fill=REGION),nudge_x = .2,nudge_y = .2,
                min.segment.length = .2, size = 3, alpha = 0.8)+
                scale_size_continuous(name = "Households (K)", breaks =
                          calculate_class(ca_geo_sub$TOTHU_CY_1,5,10000),
                labels=formatC(calculate_class(ca_geo_sub$TOTHU_CY_1,5,10000)/1000,
                big.mark=",")
                )+
              ggrepel::geom_label_repel(
                aes(label = bold_label,fill=REGION),fontface = 'bold',nudge_x = .2,nudge_y = .2,
                min.segment.length = .2, size = 3, alpha = 0.8)+
                # scale_color_identity() +
              theme(legend.position = "bottom")+
              guides(size=guide_legend(nrow = 2),
                     fill=guide_legend(nrow = 2))+
              scale_fill_manual(name = 'Region',
                                values=reg_hex_code,
                                breaks = c('Bay Area','Central Region','North Central Region','Northern Region'))+
                scale_x_continuous("Average number of extreme (24+ hrs) power disruptions a customer experienced in 2019",breaks = c(0,calculate_class(ca_geo_sub$avg_outage_customer,4,0.5)),
                                   labels = format(c(0,calculate_class(ca_geo_sub$avg_outage_customer,4,0.5)),
                                                   big.mark = ",")) +
                scale_y_continuous("% low-income population not served by HCP",
                                   breaks = calculate_class(ca_geo_sub$pct_low_income,4,5),
                                   labels = calculate_class(ca_geo_sub$pct_low_income,4,5))

scatterplot
```

- Save the figure as a PDF

```{r}
ggsave(
  here("plot", "06_scatter_extreme_poweroutage_low_income_nodist.png"),
  plot = scatterplot,
  width = 7,
  height = 5,
  # dpi = 200,
  scale = 1
)
```
